---
title: "data-carpentry-script"
output:
  html_document:
    number_sections: no
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to R

# The R syntax

# Starting with data

Download metadata file from GitHub:
```{r}
download.file("https://raw.githubusercontent.com/datacarpentry/R-genomics/gh-pages/data/Ecoli_metadata.csv", "data/Ecoli_metadata.csv")
```


# Using data in data frames

Next we want to learn how to process data that we have stored in a dataframe.

Important things to learn:

* Extract values from a vector
* Extract rows and columns from a dataframe

## Extracting values with indexing and sequences

Columns in a dataframe are vectors. Hence, we start with vectors:

Make a simple vector:

```{r}
ten_letters <- c("a","b","c","d","e","f","g","h","i","j")
```

Each value in the vector has an index. In R, indices start from 0.

We can extract a value by using the square brackets with an index:

```{r}
ten_letters[2]
```

To extract multiple values at once you can use:

* a vector of indices
* a sequence of indices. A regular sequence can be written by using the colon operator.

```{r}
ten_letters[c(1,7)]

ten_letters[3:6]

ten_letters[10:1]

ten_letters[c(2, 8:10)]
```

In case you want to extract every second value, you can create a sequence of indices using the function seq().

```{r}
seq(1, 10, by = 2)

seq(20, 4, by = -3)
```

### Exercise:

Extract every second value from the vector ten_letters:

```{r}
ten_letters[seq(1, 10, by = 2)]
```

## Extracting values, rows, and colums from dataframes

First let us load the Ecoli metadata again:

```{r}
metadata <- read.csv('data/Ecoli_metadata.csv')
```


Since dataframes are 2 dimensional (rows and columns), we need to provide indices for both dimensions. Hence the format is: dataframe[row_index, column_index]

```{r}
metadata[1, 2]   # 1st element in the 2nd column 
metadata[1, 6]   # 1st element in the 6th column
metadata[1:3, 7] # First three elements in the 7th column
metadata[3, ]    # 3rd element for all columns
metadata[, 7]    # Entire 7th column
```

Since columns can move in your dataset, you can also use the name of the column instead of the index:

```{r}
metadata[ , "strain"]
metadata[ , c("strain", "clade")]
metadata[4:7, c("strain", "clade")]
```

To extract a specific column you can also use the dollar sign operator:

dataframe$column_name

```{r}
metadata$strain
metadata$strain[2]
```




# Data manipulation using dplyr

To make data manipulation easier, people have create the package dplyr.

A package is basically a collection of functions.

To use the functions inside a package you need to install the package on your computer. We do this the following way:

```{r eval=FALSE, include=FALSE}
install.packages("dplyr") ## install
```

After the package was installed, you need to load the package.

Packages always need to be loaded at the beginning of a new R session:

```{r}
library("dplyr")          ## load
```

Dplyr is part of a collection of packages that can help you with data manipulation, visualization and analysis. Look at the following website:

https://www.tidyverse.org/

## Selecting columns and filtering rows

How to use the following dplyr functions:

* select()
* filter()
* Using pipes %>%
* mutate()
* group_by()
* summarize()

### Select()

We can use the function select() to choose columns in a dataframe.

Usage: select(.data, ...)

```{r}
select(metadata, sample, clade, cit, genome_size)
```

We can also un-select columns:

```{r}
select(metadata, -run, -genome_size)
```

### Filter()

To manipulate rows we can use the function filter():

```{r}
filter(metadata, cit == "plus") #filter for rows with samples that contain citrate using mutants

filter(metadata, cit != "plus") #exclude rows with samples that contain citrate using mutants
```

### Pipes

You can chain functions together by using the pipe operator (%>%). This operator comes from the package called magrittr, which was installed as part of the package dplyr.

```{r}
metadata %>%
  filter(cit == "plus") %>%
  select(sample, generation, clade)
```

Note that you don't need to include the data source in each function now.

You can assign the processed data to a new dataframe this way:

```{r}
meta_citplus <- metadata %>%
  filter(cit == "plus") %>%
  select(sample, generation, clade)

meta_citplus
```

#### Challenge

Using pipes, subset the data to include rows where the clade is ‘Cit+’. Retain columns sample, cit, and genome_size.

```{r}
metadata %>%
  filter(clade == "Cit+") %>%
  select(sample, cit, genome_size)
```


### Mutate()

Use the function mutate() when you want to calculate values from existing columns and store those in a new column:

```{r}
metadata %>%
  mutate(genome_bp = genome_size *1e6)
```

You can add the function head() using the pipe operator to view the top rows:

```{r}
metadata %>%
  mutate(genome_bp = genome_size *1e6) %>%
  head
```

You can remove the row countaining NA in clade the following way:

```{r}
metadata %>%
  mutate(genome_bp = genome_size *1e6) %>%
  filter(!is.na(clade)) %>%
  head
```

### group_by() and summarize()

Think about how you want to split your data to make a summary.

The group_by() function helps you to split your data.

The summarize() function helps you to create a summary of the groups:

```{r}
metadata %>%
  group_by(cit) %>%
  summarize(n())
```

In the summarize() function we used the function n() to count the observations. But we can also use functions like mean(), median(), min(), and max() to create summaries.

Missing values such as NA will interfere with certain functions. Therefore, it is best to tell summarize functions to ignore NAs:

```{r}
metadata %>%
  group_by(cit) %>%
  summarize(mean_size = mean(genome_size, na.rm = TRUE))
```

Group by multiple columns:

```{r paged.print=FALSE}
metadata %>%
  group_by(cit, clade) %>%
  summarize(mean_size = mean(genome_size, na.rm = TRUE))
```

Remove the NA in clade:

```{r}
metadata %>%
  group_by(cit, clade) %>%
  summarize(mean_size = mean(genome_size, na.rm = TRUE)) %>%
  filter(!is.na(clade))
```

Summarize multiple variables:

```{r}
metadata %>%
  group_by(cit, clade) %>%
  summarize(mean_size = mean(genome_size, na.rm = TRUE),
            min_generation = min(generation))
```

You can find a dplyr cheatsheet here: https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf



# Data Visualization

To visualize the data we can use:

* R base plots
* ggplot that is part of the tidyverse

## R base plots:

### Scatterplot

Use the function plot() to create scatterplots:

```{r}
plot(metadata$genome_size)
```

We can change the appearance of the points using the "pch"" argument:

```{r}
plot(metadata$genome_size, pch=8)
```

We can also include a title using the "main"" argument:

```{r}
plot(metadata$genome_size, pch=8, main="Scatter plot of genome sizes")
```

### Histogram

If we want to get a frequency distribution of the genome sizes we can use the hist() function:

```{r}
hist(metadata$genome_size)

hist(metadata$genome_size, xlab = "Genome size", main = "") #Rename the x-axis, remove title
```

### Boxplot

If we want to show the genome size depending on the citrate metabolism capability we could use the boxplot() function:

boxplot(y~x, data = dataframe_name)

```{r}
boxplot(genome_size ~ cit, data = metadata)
```

Improve appearance:

```{r}
boxplot(genome_size ~ cit, data = metadata,  col=c("pink","purple", "darkgrey"),
        main="Average expression differences between celltypes", ylab="Expression")
```


## Advanced figures (ggplot2)

The package ggplot2() was created to provide functions that help users to create figures in a layered approach with easier manipulation of appearance.

First we need to install the package ggplot2():

```{r eval=FALSE, include=FALSE}
install.packages("ggplot2") #install
```

Load ggplot2() package:

```{r}
library(ggplot2)
```

You will find more information about ggplot2() here: https://ggplot2.tidyverse.org/

This is the basic/minimal structure of a ggplot:

ggplot(data = <dataframe_name>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS))
  
Similar to the R base plots, there're different GEOM_FUNCTIONS for different types of plots:

* geom_point() for scatterplot
* geom_histogram() for histograms
* geom_boxplot() for boxplots

Let's build a scatterplot from metadata in layers.

First, add the data to ggplot:

```{r}
ggplot(metadata)
```

Now we need to add what type of plot we want to make. In ggplot2() the plus operator "+" is used to connect different layers. Please don't confuse with with the pipe operator %>% used previously.

```{r eval=FALSE, include=FALSE}
ggplot(metadata) +
  geom_point()
```

Lastly, we need to add information on what data we want to show. This is done in the "aesthetics" = aes():

```{r}
ggplot(metadata) +
  geom_point(aes(x = sample, y= genome_size))
```

Additionally to the x and y coordinates, we can add the following to the aesthetics:

* color (similar to line color)
* fill
* linetype (lty)
* size

First let's modify:

* color so it changes with generation
* shape so it changes with citrate usage status
* size to increase the shape size

```{r}
ggplot(metadata) +
  geom_point(aes(x = sample, y= genome_size, color = generation, shape = cit), size = rel(3.0))
```

The lables are hard to read on the x-axis.

We can change the label direction to include another layer called "theme":

```{r}
ggplot(metadata) +
  geom_point(aes(x = sample, y= genome_size, color = generation, shape = cit), size = rel(3.0)) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

### Histogram:

We can make a histogram by using the function geom_histogram(), instead of geom_point().

```{r}
ggplot(metadata) +
  geom_histogram(aes(x = genome_size))
```

As you see I get a warning to adjust the binwidth. So I choose a different binwidth:

```{r}
ggplot(metadata) +
  geom_histogram(aes(x = genome_size), binwidth=0.05)
```

### Boxplot:

```{r}
ggplot(metadata) +
  geom_boxplot(aes(x = cit, y = genome_size, fill = cit)) +
  ggtitle('Boxplot of genome size by citrate mutant type') +
  xlab('citrate mutant') +
  ylab('genome size') +
  theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=45, hjust=1),
          axis.title = element_text(size = rel(1.5)),
          axis.text = element_text(size = rel(1.25)))
```

### Save figure to pdf file

```{r}
pdf("boxplot.pdf")

ggplot(metadata) +
  geom_boxplot(aes(x = cit, y = genome_size, fill = cit)) +
  ggtitle('Boxplot of genome size by citrate mutant type') +
  xlab('citrate mutant') +
  ylab('genome size') +
  theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=45, hjust=1),
          axis.title = element_text(size = rel(1.5)),
          axis.text = element_text(size = rel(1.25)))

dev.off()
```

