---
title: "Intro R workshop"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is just text. I am not writing code yet.  

```{r}
1 + 100
```

More math in R. Uses order of operations. 

```{r}
3 + 5 * 2

(3 + 5) * 2


3 + 5 * (2 ^ 2) # ^ means to the power of
```

```{r}
2 / 10000

5e3
```

Built in math functions:

```{r}
sin(1) # trig function 

log(1) 

?log
```

Make comparisons in R. 

```{r}
#Equality 

1 == 1

#inequality  

1 != 2 #1 is not equal to 2

#less than/greater than

1 < 2 

2 > 3

2 >= 3
```

Assigning variables  

```{r}
x <- 1/40

x

log(x)

x <- 100

y <- x * 2
```

R is vectorized 

```{r}
a <- 1:5

a * 2
```

List environment

```{r}
ls()
```

Installing packages  

```{r}
install.packages("ggplot2")

library(ggplot2)
```

Challenge 

```{r}
mass <- 47.5

age <- 122 

mass <- mass * 2.3  

age <- age - 20 



a <- 4
rm(a)

```

Write code to ask if mass is larger than age. 

```{r}
mass > age
```

Data structures.  

```{r}
cats <- data.frame(coat= c("calico", "black", "tabby"), 
                   weight= c(2.1, 5.0, 3.2),
                   likes_string=c(1, 0, 1))
```

Write cat data to computer  

```{r}
write.csv(x=cats, file = "data/feline-data.csv", row.names=FALSE)

cats <- read.csv(file="data/feline-data.csv")

cats
```
Exploring data  

```{r}
str(cats)

cats$coat

cats$weight + 2
```

```{r}
paste("My cat is", cats$coat)

#cats$weight + cats$coat
```

Data types  

double, integer, complex, logical, character  

```{r}
str(cats)

typeof(cats$weight)

typeof(TRUE)
```

Learning about vectors by adding a column.  

```{r}
#vector of integers 

age <- c(2 , 3, 5)

#combine data frame with our new column we used cbind()

cbind(cats, age)


```

```{r}
#age <- c(2, 3, 4, 5)

#cbind(cats,age) wrong number of rows 
```
```{r}
dim(cats)

nrow(cats)

length(age)

nrow(cats)==length(age)
```

New column in data. 
human_age is cats$age multiplied by 7 

```{r}

cats <- cbind(cats, age)
human_age <- cats$age * 7

cats <- cbind(cats, human_age)
cats
```
```{r}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", destfile = "data/gapminder_data.csv")

gapminder <- read.csv("data/gapminder_data.csv", stringsAsFactors = TRUE)
```

Explore the data:  

```{r}
str(gapminder)

summary(gapminder$country)

summary(gapminder$gdpPercap)
```


Making ggplots  

```{r}
library(ggplot2)

ggplot(data=gapminder, mapping=aes(x=gdpPercap,y=lifeExp)) + geom_point()

#modify the plot to make a scatter plot how life expectancy has change over time (year)

ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp)) +geom_point()
```
Map a variable to a color  

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, color=continent)) + geom_point()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()+
  geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=.5) + scale_x_log10()
```
Add a linear model relationship 

```{r}
plot_test<-ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.5) + scale_x_log10() + geom_smooth(method="lm", color="red", size=2)

plot_test
```

GGplot challenge 

remake the plot above. Change the color of the points to orange. make the points size of 3. change the LM line to 1.5 size.  

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.5, color="orange", size=3) + scale_x_log10() + geom_smooth(method="lm", color="red", size=1.5) + labs(title="Life expectancy by gdp", x="GDP per capita", y="Life Exprectancy")
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp, color=continent)) + geom_point(alpha=.5) + scale_x_log10() +geom_smooth(method="lm", size=1.5)
```
Facet wrap plots by continent  

```{r}
life_ex_plot=ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp)) + 
  geom_point() +
  facet_wrap( ~ continent) +
  theme(axis.text.x=element_text(angle=45))

life_ex_plot
```

Save our ggplot

```{r}
ggsave(filename="lifeExp.png", plot=life_ex_plot, width=12, height=10, dpi= 300, units="cm")
```


GGPlot challenge 2: 
 plot lifeexp (y) by the continent on the X, colored by continent. Make boxplots. facet by year. 
 
```{r}
ggplot(data=gapminder, mapping=aes(x=continent, y=lifeExp, fill=continent)) +
  geom_boxplot()+ facet_wrap(~year) + theme(axis.title.x=element_blank(), axis.text.x = element_blank(), axis.ticks.x=element_blank(), legend.position = "bottom")
```
Manipulate gapminder data with dplyr.  

```{r}
#install.packages("dplyr")
#library("dplyr")

install.packages("tidyverse")
library(tidyverse)

#install.packages("rlang")

```

Select columns with dplyr  

```{r}
year_country_gdp <- gapminder %>% select(year, country, gdpPercap)

str(year_country_gdp)
```

Filter rows with filter. 

```{r}
year_country_gdp_euro <- gapminder %>% 
  filter(continent=="Europe") %>%
  select(year, country, gdpPercap)

str(year_country_gdp_euro)
```

Table summarizing mean gdp per cap, by country  

```{r}
gdp_bycont <- gapminder %>% group_by(continent) %>% summarize(mean_gdpPerCap=mean(gdpPercap))
```

